%%%% -*- Mode: LaTeX -*-

%%%% infinities.tex

\documentclass[../Float-Infinities-Nan.tex]{subfiles}
\begin{document}

\DDictionaryItem{Constant Variables\\
  \code{long-float-positive-infinity},
  \code{long-float-negative-infinity},\\
  \code{double-float-positive-infinity},
  \code{double-float-negative-infinity},\\
  \code{single-float-positive-infinity},
  \code{single-float-negative-infinity},\\
  \code{short-float-positive-infinity},
  \code{short-float-negative-infinity},\\
  \code{infL},
  \code{infD},
  \code{infF},
  \code{infS}
}
\index{I!Infinities!\code{long-float-positive-infinity}}
\index{I!Infinities!\code{long-float-negative-infinity}}
\index{I!Infinities!\code{double-float-positive-infinity}}
\index{I!Infinities!\code{double-float-negative-infinity}}
\index{I!Infinities!\code{single-float-positive-infinity}}
\index{I!Infinities!\code{single-float-negative-infinity}}
\index{I!Infinities!\code{short-float-positive-infinity}}
\index{I!Infinities!\code{short-float-negative-infinity}}
\index{I!Infinities!\code{infL}}
\index{I!Infinities!\code{infD}}
\index{I!Infinities!\code{infF}}
\index{I!Infinities!\code{infS}}  
% \index{I!Infinities!\code{+infL0}}, 
% \index{I!Infinities!\code{-infL0}},
% \index{I!Infinities!\code{+infD0}}, 
% \index{I!Infinities!\code{-infD0}},
% \index{I!Infinities!\code{+infF0}}, 
% \index{I!Infinities!\code{-infF0}},
% \index{I!Infinities!\code{+infS0}}, 
% \index{I!Infinities!\code{-infS0}},

\DValues{}

The value of each of these constants is
\emph{implementation-dependent}.

\DDescription{}

The value of each of these constants must conform with the format
(i.e., the floating point type) codified in the name.

\DExamples{}

\begin{alltt}
SBCL> \codeprompt{single-float-positive-infinity}
\textit{#.SB-EXT:SINGLE-FLOAT-POSITIVE-INFINITY}
\end{alltt}

\begin{alltt}
LW> \codeprompt{single-float-positive-infinity}
\textit{+1F++0} \textcolor{red}{#| +1F++0 is single-float plus-infinity |#}
\end{alltt}

\begin{alltt}
LW> \codeprompt{-infD0}
\textit{-1D++0} \textcolor{red}{#| -1D++0 is double-float plus-infinity |#}
\end{alltt}

\begin{alltt}
LW> \codeprompt{-infL0}
\textit{-1D++0} \textcolor{red}{#| -1D++0 is double-float plus-infinity |#}
\textcolor{red}{;;; Note that in this case the widest float representation
;;; available is DOUBLE-FLOAT.}
\end{alltt}


\DNotes{}

The ``short'' names and the ``long ones'' are completely equivalent.
A possible implementation would be the following:
\begin{alltt}
(define-symbol-macro -infD0 double-float-negative-infinity)
\end{alltt}

\end{document}